---
title: "Using RKeOps"
output:
  rmarkdown::html_vignette:
    toc: true
  pdf_document:
    toc: true
    number_sections: yes
author: "Ghislain Durif"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Using RKeOps}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  progress = TRUE,
  warning = FALSE,
  eval = FALSE
)
```

RKeOps is a R front-end for the KeOps C++/Cuda library. It provides standard functions that can be used in any R code.<br><br>

Thanks to RKeOps, you can use **GPU computing directly inside R** without the cost of developing a specific CUDA implementation of your custom mathematical operators.

# Installing RKeOps

## Requirements

* R (tested with R >= 4.2)
* C++ compiler (g++ >=7 or clang) for CPU computing or CUDA compiler (nvcc >=10) and CUDA libs for GPU computing
* Python (>= 3.10)

**Important:** Python is a requirement as an intern machinery for the package to work but you will not need to create nor manipulate Python codes to use the RKeOps package.

**Disclaimer:** KeOps (including RKeOps) is not functional on Windows, it was only tested on Linux and MacOS.

## Install from CRAN

> **Note:** RKeOps is avaible on CRAN but only for UNIX environment (GNU/Linux and MacOS) and not for Windows.

```{r install}
install.packages("rkeops")
```

<!-- RKeOps >=2 is not available on CRAN yet, please refer to the next section. -->

## Install development version

```{r install_github}
install.packages("remotes")
remotes::install_github("getkeops/keops", subdir = "rkeops")
```

## Setup and requirement install

**To be done once after installation:**

```{r install_requirements}
# load rkeops
library(rkeops)
# create a dedicated Python environment with reticulate (to be done only once)
reticulate::virtualenv_create("rkeops")
# activate the dedicated Python environment
reticulate::use_virtualenv(virtualenv = "rkeops", required = TRUE)
# install rkeops requirements (to be done only once)
install_rkeops()
```

### Note on Python requirement

RKeOps now uses PyKeOps Python package under the hood thanks to the [`reticulate`](https://rstudio.github.io/reticulate/) R package that provides an “R Interface to Python”.

We recommend to use a dedicated Python environment (through `reticulate`) to install and use RKeOps Python dependencies (c.f. [previously](#setup-and-requirement-install)).

> **Note:** To get more information about **managing** which **version of Python** you are using, you can refer to `reticulate` [documentation](https://rstudio.github.io/reticulate/articles/versions.html) about “Python Version Configuration”. In particular, if you are a Miniconda/Anaconda Python distribution user, you can either user a Python virtual environment (c.f. above) or a **Conda environment** with the same results. Please refer to the `reticulate` [documentation](https://rstudio.github.io/reticulate/articles/versions.html#providing-hints) in this case.

---

# How to use RKeOps

Load RKeOps in R:
```{r setup}
# load rkeops
library(rkeops)
# activate the dedicated Python environment
reticulate::use_virtualenv(virtualenv = "rkeops", required = TRUE)
```

**Note**: If you are using a Python environment (c.f. [previously](#note-on-python-requirement)), you should always activate it after loading RKeOps.

You can verify that every thing is ok:
```{r check}
check_rkeops()
```

RKeOps allows to define and compile new operators that run computations on GPU.

## LazyTensors

Lazy evaluation allows to write intermediate computations as symbolic operations that are not directly evaluated. The real evaluation is only made on final computation.

To do so, you can use `LazyTensors`. These are objects wrapped around R matrices or vectors used to create symbolic formulas for the `KeOps` reduction operations. A typical use case is the following: 

Let us say that we want to compute (for all $j=1,\dots,N$):

$$
a_j = \sum_{i=1}^{M} \exp\left(-\frac{\Vert \mathbf x_i - \mathbf y_j\Vert^2}{s^2}\right)
$$

with

* parameter: $s \in\mathbb R$<br>

* $i$-indexed variables $\mathbf X = [\mathbf x_i]_{i=1,...,M} \in\mathbb R^{M\times 3}$<br>

* $j$-indexed variables $\mathbf Y = [\mathbf y_j]_{j=1,...,N} \in\mathbb R^{N\times 3}$<br><br>

The associated code would be:

```{r}
# to run computation on CPU (default mode)
rkeops_use_cpu()
# OR
# to run computations on GPU (to be used only if relevant)
rkeops_use_gpu()

# Data
M <- 10000
N <- 15000
x <- matrix(runif(N * 3), nrow = M, ncol = 3) # arbitrary R matrix representing 
                                              # 10000 data points in R^3
y <- matrix(runif(M * 3), nrow = N, ncol = 3) # arbitrary R matrix representing 
                                              # 15000 data points in R^3
s <- 0.1                                      # scale parameter

# Turn our Tensors into KeOps symbolic variables:
x_i <- LazyTensor(x, "i")     # symbolic object representing an arbitrary row of x, 
                              # indexed by the letter "i"
y_j <- LazyTensor(y, "j")     # symbolic object representing an arbitrary row of y, 
                              # indexed by the letter "j"

# Perform large-scale computations, without memory overflows:
D_ij <- sum((x_i - y_j)^2)    # symbolic matrix of pairwise squared distances, 
                              # with 10000 rows and 15000 columns

K_ij <- exp(- D_ij / s^2)     # symbolic matrix, 10000 rows and 15000 columns

# D_ij and K_ij are only symbolic at that point, no computation is done

# Computing the result without storing D_ij and K_ij:
a_j <- sum(K_ij, index = "i") # actual R matrix (in fact a row vector of 
                              # length 15000 here)
                              # containing the column sums of K_ij
                              # (i.e. the sums over the "i" index, for each 
                              # "j" index)
```

**More in the dedicated "RKeOps LazyTensor" [article](https://www.kernel-operations.io/rkeops/articles/LazyTensor_rkeops.html)** or in the corresponding vignette:
```{r eval=FALSE}
vignette("LazyTensor_rkeops", package = "rkeops")
```

## Generic reductions

With RKeOps, you can also directly define new operator with plain mathematical formula written as a character string:

```{r}
# implementation of a convolution with a Gaussian kernel
formula = "Sum_Reduction(Exp(-s * SqNorm2(x - y)) * b, 0)"
# input arguments
args = c("x = Vi(3)",      # vector indexed by i (of dim 3)
         "y = Vj(3)",      # vector indexed by j (of dim 3)
         "b = Vj(6)",      # vector indexed by j (of dim 6)
         "s = Pm(1)")      # parameter (scalar)
# compilation
op <- keops_kernel(formula, args)
# data and parameter values
nx <- 100
ny <- 150
X <- matrix(runif(nx*3), nrow=nx)   # matrix 100 x 3
Y <- matrix(runif(ny*3), nrow=ny)   # matrix 150 x 3
B <- matrix(runif(ny*6), nrow=ny)   # matrix 150 x 6
s <- 0.2

# to run computation on CPU (default mode)
use_cpu()
# to run computations on GPU (to be used only if relevant)
use_gpu()

# computation (order of the input arguments should be similar to `args`)
res <- op(list(X, Y, B, s))
```

The different elements (formula, arguments, compilation, computation) in the previous example will be detailed in the next sections.

### Formula

To use RKeOps and define new operators, you need to write the corresponding _formula_ which is a text string defining a composition of mathematical operations. It should be characterized by two elements:

1. a composition of generic functions applied to some input matrices, whose one of their dimensions is either indexed by $i=1,...,M$ or $j=1,...,N$

2. a reduction over indexes $i=1,...,M$ (row-wise) or $j=1,...,N$ (column-wise) of the $M \times N$ matrix whose entries are defined by 1.

RKeOps implements a wide range of mathematical operators and reduction: please refers to
<https://www.kernel-operations.io/keops/api/math-operations.html> for more details.<br><br>

**Example:** We want to implement the following kernel-based reduction (convolution with a Gaussian kernel): $$\sum_{j=1}^{N} \exp\Big(-\sigma || \mathbf x_i - \mathbf y_j ||_2^{\,2}\Big)\,\mathbf b_j$$
with

* parameter: $\sigma\in\mathbb R$<br>

* $i$-indexed variables $[\mathbf x_i]_{i=1,...,M} \in\mathbb R^{M\times 3}$<br>

* $j$-indexed variables $[\mathbf y_j]_{j=1,...,N} \in\mathbb R^{N\times 3}$ and $[\mathbf b_j]_{j=1,...,N} \in\mathbb R^{N\times 6}$<br><br>

In R, we can define the corresponding KeOps formula as a simple **text string**:
```{r formula}
formula = "Sum_Reduction(Exp(-s * SqNorm2(x - y)) * b, 0)"
```
* `SqNorm2` = squared $\ell_2$ norm
* `Exp` = exponential
* `Sum_reduction(..., 0)` = sum reduction over the dimension 0 i.e. sum on the $j$'s (1 to sum over the $i$'s)<br>

### Arguments

The formula describing your computation can take several input arguments: variables and parameters. The input variables will generally corresponds to rows or columns of your data matrices, you need to be cautious with their dimensions.

#### Input matrix

You can use two type of input matrices with RKeOps:<br>

* ones whose rows (or columns) are indexed by $i=1,...,M$ such as $\mathbf X = [x_{ik}]_{M \times D}$<br>

* others whose rows (or columns) are indexed by $j=1,...,N$ such as $\mathbf Y = [y_{ik'}]_{N \times D'}$<br><br>

The dimensions over indexes $i$ or $j$ are called the **outer dimensions** (i.e. $M$ or $N$). The other dimensions (i.e. $D$ or $D'$) are called the **inner dimensions**. These terms refer to the contiguity of the data in memory:<br>

* **Outer dimensions** $M$ and $N$ (over indexes $i$ and $j$ respectively) can be **very large**, even too large for GPU memory.<br>

* **Inner dimensions** $D$ and $D'$ should be **small** enough to fit in GPU memory, in particular to ensure data colocality and avoid useless memory transfers. Corresponding columns (or rows) should be contiguous in memory (this point is handled for you in RKeOps, see this [section](#data-storage-orientation)).<br>

> **Note 1:** The outer dimension can correspond to the rows or the columns of the input matrices (and vice-versa for the inner dimension). The optimal orientation of input matrices is discussed in this [section](#data-storage-orientation) .

> **_Note 2:_** All matrices indexed by $i$ should have the same outer dimension $M$ over $i$, same for all matrices indexed by $j$ (outer dimension $N$). **Only the inner dimensions $D$ and $D'$ should be known for the compilation of your operators.** The respective outer dimensions $M$ and $N$ are discovered at runtime (and can change from one run to another).<br>

#### Notations

Input arguments of the formula are defined by using keywords, they can be of different types:

| keyword | meaning                 |
|:--------|:------------------------|
| `Vi`    | variable indexed by `i` |
| `Vj`    | variable indexed by `j` |
| `Pm`    | parameter               |

You should provide a vector of text string specifying the name and the type of all arguments in your formula.

Each keyword takes as parameter the inner dimension of the corresponding object. For instance, to define an input variable indexed by $i$ corresponding to a $D$-dimensional vector, you can use `"Vi(D)"`, same for a $D$-dimensional variable indexed by $j$ being `"Vj(D)"` or a $D$-dimensional parameter `"Pm(D)"`.<br><br>

The vector of arguments should be
```{r template_args}
args = c("<name1>=<type1>(dim1)", "<name2>=<type2>(dim2)", "<nameX>=<typeX>(dimX)")
```
where

* `<nameX>` is the name
* `<type1>` is the type (among `Vi`, `Vj` or `Pm`)
* `<dimX>` is the **inner dimension**

of the `X`$^\text{th}$ variable in the formula.<br><br>

> **Important:** The names should correspond to the ones used in the formula. The input parameter order will be the one used when calling the compiled operator.

**Example:** We define the corresponding arguments of the previous [formula](#formula), i.e. parameters or variables indexed by $i$ or $j$ with their corresponding inner dimensions:
```{r args}
args = c("x = Vi(3)",      # vector indexed by i (of dim 3)
         "y = Vj(3)",      # vector indexed by j (of dim 3)
         "b = Vj(6)",      # vector indexed by j (of dim 6)
         "s = Pm(1)")      # parameter (scalar)
```



### Creating a new operator

By using the function `keops_kernel`, based on the formula and its arguments that we previously defined, we can compile and load into R the corresponding operator:
```{r compile}
# compilation
op <- keops_kernel(formula, args)
```

Calling `keops_kernel(formula, args)` returns a function that can be later used to run computations on your data with your value of parameters. You should only be cautious with the similarity of each argument inner dimension.<br><br>

The returned function (here `op`) expects a list of input values in the order specified in the vector `args`.

The result of compilation (shared library file) is stored on the system and will be reused when calling again the function `keops_kernel` on the same formula with the same arguments and the same conditions (e.g. precision), to avoid useless recompilation.

### Run computations

We generate data with inner dimensions (number of columns) corresponding to each arguments expected by the operator `op`. The function `op` takes in input a list of input arguments. If the list if named, `op` checks the association between the supplied names and the names of the formula arguments. In this case only, it can also correct the order of the input list to match the expected order of arguments.

```{r run}
# data and parameter values
nx <- 100
ny <- 150
X <- matrix(runif(nx*3), nrow=nx)   # matrix 100 x 3
Y <- matrix(runif(ny*3), nrow=ny)   # matrix 150 x 3
B <- matrix(runif(ny*6), nrow=ny)   # matrix 150 x 6
s <- 0.2

# to run computation on CPU (default mode)
rkeops_use_cpu()
# OR
# to run computations on GPU (to be used only if relevant)
rkeops_use_gpu()

# computation (order of the input arguments should be similar to `args`)
res <- op(list(x, y, beta, s))
```

### Computing gradients

You can define gradients directly in the formula, e.g.
```{r grad}
# defining a formula with a Gradient
formula <- "Grad(Sum_Reduction(SqNorm2(x-y), 0), x, eta)"
args <- c("x=Vi(0,3)", "y=Vj(1,3)", "eta=Vi(2,1)")
# compiling the corresponding operator
op <- keops_kernel(formula, args)

# data
nx <- 100
ny <- 150
x <- matrix(runif(nx*3), nrow=nx, ncol=3)     # matrix 100 x 3
y <- matrix(runif(ny*3), nrow=ny, ncol=3)     # matrix 150 x 3
eta <- matrix(runif(nx*1), nrow=nx, ncol=1)   # matrix 100 x 1

# computation
input <- list(x, y, eta)
res <- op(input)
```
where `eta` is the new variable at which the gradient is computed, its dimension should correspond to the output dimension of the operation inside the gradient (here `SqNorm2(x-y)` is of dimension 1).

You can also use the function `keops_grad` to compute partial derivative for existing KeOps operators.
```{r keops_grad}
# defining an operator (reduction on squared distance)
formula <- "Sum_Reduction(SqNorm2(x-y), 0)"
args <- c("x=Vi(0,3)", "y=Vj(1,3)")
op <- keops_kernel(formula, args)
# defining its gradient regarding x
grad_op <- keops_grad(op, var="x")

# data
nx <- 100
ny <- 150
x <- matrix(runif(nx*3), nrow=nx, ncol=3)     # matrix 100 x 3
y <- matrix(runif(ny*3), nrow=ny, ncol=3)     # matrix 150 x 3
eta <- matrix(runif(nx*1), nrow=nx, ncol=1)   # matrix 100 x 1

# computation
input <- list(x, y, eta)
res <- grad_op(input)
```

**Note:** when defining a gradient, the operator created by `keops_grad`requires an additional variable, of the same type (`Vi`, `Vj`, `Pm`) as the variable with respect to which the differentiation is made (here `x` hence `Vi`), and whose inner dimension corresponds to the output dimension of the derived formula (here `SqNorm2(x-y)` is a real-valued function, hence dimension 1). In addition, (if `Vi`, `Vj` variable), its outer dimension should corresponds to the outer dimension of the variable regarding which the gradient is taken (here `x`).


## RKeOps options

RKeOps behavior is driven by specific options in `R` global options scope. Such options are set up when loading RKeOps (i.e. by calling `library(rkeops)`).

You can get the current values of RKeOps options with
```{r get_options}
get_rkeops_options()
```

To (re)set RKeOps options to default values, run:
```{r reset_options}
set_rkeops_options()
```

You can either use `set_rkeops_options()` to modify a specific options (c.f. `?set_rkeops_options`) or you can use helper functions, c.f. below.

### Choosing CPU or GPU computing at runtime

By default, RKeOps runs computations on CPU (even for GPU-compiled operators). To enable GPU computing, you can run (before calling your operator):
```{r use_gpu}
rkeops_use_gpu()
```
You can also specify the GPU id that you want to use, e.g. `rkeops_use_gpu(device=0)`
to use GPU 0 (default) for instance.

To deactivate GPU computations, you can run `rkeops_use_cpu()`.

> In CPU mode, you can control the number of CPU cores used by RKeOps for computations, e.g. with `rkeops_use_cpu(ncore = 2)` to run on 2 cores.

### Computation precision

By default, RKeOps uses  32bits float precision for computations. Since R only considers 64bits floating point numbers, if you want to use float 32bits, input data and output results will be casted befors and after computations respectively in your RKeOps operator. If your application requires to use 64bits float (double) precision, keep in mind that you will suffer a performance loss (potentially not an issue on high-end GPUs). In any case, compensated summation reduction is available in KeOps to correct for 32bits floating point arithmetic errors (see `sum_scheme` argument for `keops_kernel()` function, c.f. `?keops_kernel`).

You use the following functions to choose between 32bits float and 64bits float precision (respectively):

```{r precision}
rkeops_use_float32()
rkeops_use_float64()
```

### Verbosity

You can use the following functions to respectively enable and disable compilation verbosity:

```{r verbosity}
rkeops_enable_verbosity()
rkeops_disable_verbosity()
```


## Data storage orientation

In R, matrices are stored using a column-major order, meaning that a $M \times D$ matrix is stored in memory as a succession of $D$ vectors of length $M$ representing each of its columns. A consequence is that two successive entries of a column are contiguous in memory, but two successive entries of a row are separated by $M$ elements. See this [page](https://en.wikipedia.org/wiki/Row-_and_column-major_order) for more details.<br><br>

**_Important:_** In machine learning and statistics, we generally use data matrices where each sample/observation/individual is a row, i.e. matrices where the outer dimensions correspond to rows, e.g. $\mathbf X = [x_{ik}]_{M \times D}$, $\mathbf Y = [y_{ik'}]_{N \times D'}$. This is the default using case of RKeOps. RKeOps will then automatically process the input data to enforce colocality along the inner dimension.

If you want to use data where the inner dimension corresponds to rows of your matrices, i.e. $\mathbf X^{t} = [x_{ki}]_{D \times M}$ or $\mathbf Y^{t} = [y_{k'i}]_{D' \times N}$, you just need to specify the input parameter `inner_dim="row"` when calling your operator.

Example:
```{r storage_order}
# standard column reduction of a matrix product
op <- keops_kernel(formula = "Sum_Reduction((x|y), 1)",
                   args = c("x=Vi(3)", "y=Vj(3)"))

# data (inner dimension = columns)
nx <- 10
ny <- 15
# x_i = rows of the matrix X
X <- matrix(runif(nx*3), nrow=nx, ncol=3)
# y_j = rows of the matrix Y
Y <- matrix(runif(ny*3), nrow=ny, ncol=3)
# computing the result (here, by default `inner_dim="col"` and columns 
# corresponds to the inner dimension)
res <- op(list(X,Y))

# data (inner dimension = rows)
nx <- 10
ny <- 15
# x_i = columns of the matrix X
X <- matrix(runif(nx*3), nrow=3, ncol=nx)
# y_j = columns of the matrix Y
Y <- matrix(runif(ny*3), nrow=3, ncol=ny)
# computing the result (we specify `inner_dim="row"` to indicate that rows
# corresponds to the inner dimension)
res <- op(list(X,Y), inner_dim="row")
```

### Compilation files and cleaning

The compilation of new operators produces shared library (or share object `.so`) files stored in a `build` sub-directory of the package installation directory, to be reused and avoid recompilation of already defined operators.

You can check where your compiled operators are stored by running `get_rkeops_build_dir()`. To clean RKeOps install and remove all shared library files, you can run `clean_rkeops()`.
