---
title: "Using RKeOps"
output: 
  rmarkdown::html_vignette:
    toc: true
  pdf_document:
    toc: true
    number_sections: yes
author: ""
date: "`r Sys.Date()`"
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{RKeOps: R bindings for KeOps}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  progress = TRUE,
  warning = FALSE
)
```

RKeOps is a R front-end for the KeOps C++/Cuda library. It provides standard functions that can be used in any R code.<br><br>

Thanks to RKeOps, you can use **GPU computing directly inside R** without the cost of developing a specific CUDA implementation of your custom mathematical operators.

# Installing RKeOps

## Requirements

* R (tested with R >= 3.5)
* Cmake (>=3.10)
* C++ compiler (g++ >=7 or clang) for CPU computing or CUDA compiler (nvcc >=10) and CUDA libs for GPU computing

**Disclaimer:** KeOps (including RKeOps) is not functional on Windows, it was only tested on Linux and MacOS.

## Install from CRAN

> **Note:** RKeOps is avaible on CRAN but only for UNIX environment (GNU/Linux and MacOS) and not for Windows.

```{r install, eval=FALSE}
install.packages("rkeops")
```

## Install from Github sources

> !! In most recent version of devtools, the `args` argument is not available anymore and it is not possible to use `devtools::install_git`. Please check next section to install from sources.

* Install directly from Github (requires `git`)
```{r install_github, eval=FALSE}
devtools::install_git("https://github.com/getkeops/keops", 
                      subdir = "rkeops", 
                      args="--recursive")
# not possible to use `devtools::intall_github()` because of the required submodule
```

## Get sources and install from local repository

* Get KeOps sources (bash command)
```bash
git clone --recurse-submodules="keops/lib/sequences" https://github.com/getkeops/keops
# or
git clone https://github.com/getkeops/keops
cd keops
git submodule update --init -- keops/lib/sequences
# other submodules are not necessary for RKeOps
```

* Install from local source in R (assuming you are in the `keops` directory)
```{r install_src, eval=FALSE}
devtools::install("rkeops")
```

---

# How to use RKeOps

Load RKeOps in R:
```{r setup}
library(rkeops)
```

RKeOps allows to define and compile new operators that run computations on GPU.

## Example

```{r full_example, eval=FALSE}
# implementation of a convolution with a Gaussian kernel
formula = "Sum_Reduction(Exp(-s * SqNorm2(x - y)) * b, 0)"
# input arguments
args = c("x = Vi(3)",      # vector indexed by i (of dim 3)
         "y = Vj(3)",      # vector indexed by j (of dim 3)
         "b = Vj(6)",      # vector indexed by j (of dim 6)
         "s = Pm(1)")      # parameter (scalar) 
# compilation
op <- keops_kernel(formula, args)
# data and parameter values
nx <- 100
ny <- 150
X <- matrix(runif(nx*3), nrow=nx)   # matrix 100 x 3
Y <- matrix(runif(ny*3), nrow=ny)   # matrix 150 x 3
B <- matrix(runif(ny*6), nrow=ny)   # matrix 150 x 6
s <- 0.2

# to run computation on CPU (default mode)
use_cpu()
# to run computations on GPU (to be used only if relevant)
use_gpu()

# computation (order of the input arguments should be similar to `args`)
res <- op(list(X, Y, B, s))
```

The different elements (formula, arguments, compilation, computation) in the previous example will be detailled in the next sections.

## Formula

To use RKeOps and define new operators, you need to write the corresponding _formula_ which is a text string defining a composition of mathematical operations. It should be characterized by two elements:

1. a composition of generic functions applied to some input matrices, whose one of their dimensions is either indexed by $i=1,...,M$ or $j=1,...,N$

2. a reduction over indexes $i=1,...,M$ (row-wise) or $j=1,...,N$ (column-wise) of the $M \times N$ matrix whose entries are defined by 1.

RKeOps implements a wide range of mathematical operators and reduction: please refers to 
<https://www.kernel-operations.io/keops/api/math-operations.html> for more details.<br><br>

**Example:** We want to implement the following kernel-based reduction (convolution with a Gaussian kernel): $$\sum_{j=1}^{N} \exp\Big(-\sigma || \mathbf x_i - \mathbf y_j ||_2^{\,2}\Big)\,\mathbf b_j$$
with

* parameter: $\sigma\in\mathbb R$<br>

* $i$-indexed variables $[\mathbf x_i]_{i=1,...,M} \in\mathbb R^{M\times 3}$<br>

* $j$-indexed variables $[\mathbf y_j]_{j=1,...,N} \in\mathbb R^{N\times 3}$ and $[\mathbf b_j]_{j=1,...,N} \in\mathbb R^{N\times 6}$<br><br>

In R, we can define the corresponding KeOps formula as a simple **text string**:
```{r formula, eval=FALSE}
formula = "Sum_Reduction(Exp(-s * SqNorm2(x - y)) * b, 0)"
```
* `SqNorm2` = squared $\ell_2$ norm
* `Exp` = exponential
* `Sum_reduction(..., 0)` = sum reduction over the dimension 0 i.e. sum on the $j$'s (1 to sum over the $i$'s)<br>

## Arguments

The formula describing your computation can take several input arguments: variables and parameters. The input variables will generally corresponds to rows or columns of your data matrices, you need to be cautious with their dimensions.

### Input matrix

You can use two type of input matrices with RKeOps:<br>

* ones whose rows (or columns) are indexed by $i=1,...,M$ such as $\mathbf X = [x_{ik}]_{M \times D}$<br>

* others whose rows (or columns) are indexed by $j=1,...,N$ such as $\mathbf Y = [y_{ik'}]_{N \times D'}$<br><br>

The dimensions over indexes $i$ or $j$ are called the **outer dimensions** (i.e. $M$ or $N$). The other dimensions (i.e. $D$ or $D'$) are called the **inner dimensions**. These terms refer to the contiguity of the data in memory:<br>

* **Outer dimensions** $M$ and $N$ (over indexes $i$ and $j$ respectively) can be **very large**, even to large for GPU memory.<br>

* **Inner dimensions** $D$ and $D'$ should be **small** enough to fit in GPU memory, in particular to ensure data colocality and avoid useless memory transfers. Corresponding columns (or rows) should be contiguous in memory (this point is handled for you in RKeOps, see this [section](#data-storage-orientation)).<br>

> **_Note 1:_** The outer dimension can correspond to the rows or the columns of the input matrices (and vice-versa for the inner dimension). The optimal orientation of input matrices is discussed in this [section](#data-storage-orientation) .

> **_Note 2:_** All matrices indexed by $i$ should have the same outer dimension $M$ over $i$, same for all matrices indexed by $j$ (outer dimension $N$). Only the inner dimensions $D$ and $D'$ should be known for the compilation of your operators. The respective outer dimensions $M$ and $N$ are set at runtime (and can change from one run to another).<br>

### Notations

Input arguments of the formula are defined by using keywords, they can be of different types: 

| keyword | meaning                 |
|:--------|:------------------------|
| `Vi`    | variable indexed by `i` |
| `Vj`    | variable indexed by `j` |
| `Pm`    | parameter               |

You should provide a vector of text string specifying the name and the type of all arguments in your formula.

Each keyword takes as parameter the inner dimension of the corresponding object. For instance, to define an input variable indexed by $i$ corresponding to a $D$-dimensional vector, you can use `"Vi(D)"`, same for a $D$-dimensional variable indexed by $j$ being `"Vj(D)"` or a $D$-dimensional parameter `"Pm(D)"`.<br><br>

The vector of arguments should be
```{r template_args, eval=FALSE}
args = c("<name1>=<type1>(dim1)", "<name2>=<type2>(dim2)", "<nameX>=<typeX>(dimX)")
```
where

* `<nameX>` is the name
* `<type1>` is the type (among `Vi`, `Vj` or `Pm`)
* `<dimX>` is the **inner dimension**

of the `X`$^\text{th}$ variable in the formula.<br><br>

> **_Important:_** The names should correspond to the ones used in the formula. The input parameter order will be the one used when calling the compiled operator.

**Example:** We define the corresponding arguments of the previous [formula](#formula), i.e. parameters or variables indexed by $i$ or $j$ with their corresponding inner dimensions:
```{r args, eval=FALSE}
args = c("x = Vi(3)",      # vector indexed by i (of dim 3)
         "y = Vj(3)",      # vector indexed by j (of dim 3)
         "b = Vj(6)",      # vector indexed by j (of dim 6)
         "s = Pm(1)")      # parameter (scalar) 
```



## Creating a new operator

By using the function `keops_kernel`, based on the formula and its arguments that we previously defined, we can compile and load into R the corresponding operator:
```{r compile, eval=FALSE}
# compilation
op <- keops_kernel(formula, args)
```

Calling `keops_kernel(formula, args)` returns a function that can be later used to run computations on your data with your value of parameters. You should only be cautious with the similarity of each argument inner dimension.<br><br>

The returned function (here `op`) expects a list of input values in the order specified in the vector `args`.

The result of compilation (shared library file) is stored on the system and will be reused when calling again the function `keops_kernel` on the same formula with the same arguments and the same conditions (e.g. precision), to avoid useless recompilation.

## Run computations

We generate data with inner dimensions (number of columns) corresponding to each arguments expected by the operator `op`. The function `op` takes in input a list of input arguments. If the list if named, `op` checks the association between the supplied names and the names of the formula arguments. In this case only, it can also correct the order of the input list to match the expected order of arguments.

```{r run, eval=FALSE}
# data and parameter values
nx <- 100
ny <- 150
X <- matrix(runif(nx*3), nrow=nx)   # matrix 100 x 3
Y <- matrix(runif(ny*3), nrow=ny)   # matrix 150 x 3
B <- matrix(runif(ny*6), nrow=ny)   # matrix 150 x 6
s <- 0.2

# to run computation on CPU (default mode)
use_cpu()
# to run computations on GPU (to be used only if relevant)
use_gpu()

# computation (order of the input arguments should be similar to `args`)
res <- op(list(x, y, beta, s))
```

## Computing gradients

You can define gradients directly in the formula, e.g.
```{r grad, eval=FALSE}
# defining a formula with a Gradient
formula <- "Grad(Sum_Reduction(SqNorm2(x-y), 0), x, eta)"
args <- c("x=Vi(0,3)", "y=Vj(1,3)", "eta=Vi(2,1)")
# compiling the corresponding operator
op <- keops_kernel(formula, args)

# data
nx <- 100
ny <- 150
x <- matrix(runif(nx*3), nrow=nx, ncol=3)     # matrix 100 x 3
y <- matrix(runif(ny*3), nrow=ny, ncol=3)     # matrix 150 x 3
eta <- matrix(runif(nx*1), nrow=nx, ncol=1)   # matrix 100 x 1

# computation
input <- list(x, y, eta)
res <- op(input)
```
where `eta` is the new variable at which the gradient is computed, its dimension should correspond to the output dimension of the operation inside the gradient (here `SqNorm2(x-y)` is of dimension 1).

You can also use the function `keops_grad` to derive existing KeOps operators.
```{r keops_grad, eval=FALSE}
# defining an operator (reduction on squared distance)
formula <- "Sum_Reduction(SqNorm2(x-y), 0)"
args <- c("x=Vi(0,3)", "y=Vj(1,3)")
op <- keops_kernel(formula, args)
# defining its gradient regarding x
grad_op <- keops_grad(op, var="x")

# data
nx <- 100
ny <- 150
x <- matrix(runif(nx*3), nrow=nx, ncol=3)     # matrix 100 x 3
y <- matrix(runif(ny*3), nrow=ny, ncol=3)     # matrix 150 x 3
eta <- matrix(runif(nx*1), nrow=nx, ncol=1)   # matrix 100 x 1

# computation
input <- list(x, y, eta)
res <- grad_op(input)
```

**Note:** when defining a gradient, the operator created by `keops_grad`requires an additional variable whose inner dimension corresponds to the output dimension of the derived formula (here `SqNorm2(x-y)` is a real-valued function, hence dimension 1) and outer dimension corresponds to the outer dimension of the variable regarding which the gradient is taken (here `x`).


## RKeOps options

RKeOps behavior is driven by specific options in `R` global options scope. Such options are set up when loading RKeOps (i.e. by calling `library(rkeops)`).

You can get the current values of RKeOps options with
```{r get_options, eval=FALSE}
get_rkeops_options()
```

To (re)set RKeOps options to default values, run:
```{r reset_options, eval=FALSE}
set_rkeops_options()
```

To set a specific option with a given value, you can do:
```{r set_option, eval=FALSE}
set_rkeops_option(option, value)
# `option` = text string, name of the option to set up
# `value` = whatever value to assign to the chosen option
```
Check `?set_rkeops_option` for more details.

### Compile options

* `use_cuda_if_possible`: by default, user-defined operators are compiled for GPU if CUDA is available (and compiled for CPU otherwise).
```{r set_cuda, eval=FALSE}
# enable compiling for GPU if available (not necessary if using default options)
compile4gpu()
# or equivalently
set_rkeops_option("use_cuda_if_possible", 1)
# disable compiling for GPU
set_rkeops_option("use_cuda_if_possible", 0)
```

* `precision`: by default, user-defined operators are compiled to use float 32bits for computations (faster than float 64bits or double, compensated sum is available to reduce errors inherent to float 32bits operations)
```{r set_precision, eval=FALSE}
set_rkeops_option("precision", "float")    # float 32bits (default)
set_rkeops_option("precision", "double")   # float 64bits
```

You can directly change the precision used in compiled operators with the 
functions `compile4float32` and `compile4float64` which respectively 
enable float 32bits precision (default) and float 64bits (or double) precision.

* other compile options (including boolean value to enable verbosity or to 
add debugging flag), see `?compile_options`


### Choosing CPU or GPU computing at runtime

By default, RKeOps runs computations on CPU (even for GPU-compiled operators). To enable GPU computing, you can run (before calling your operator):
```{r use_gpu, eval=FALSE}
use_gpu()
# see `?runtime_options` for a more advanced use of GPU inside RKeOps
```
You can also specify the GPU id that you want to use, e.g. `use_gpu(device=0)` 
to use GPU 0 (default) for instance.

To deactivate GPU computations, you can run `use_cpu()`.

> In CPU mode, you can control the number of CPU cores used by RKeOps for computations, e.g. with `use_cpu(ncore = 2)` to run on 2 cores.

### Other runtime options

* `device_id`: choose on which GPU the computations will be done, default is 0.
```{r set_device_id, eval=FALSE}
set_rkeops_option("device_id", 0)
```

**_Note_**: We recommend to handle GPU assignation outside RKeOps, for instance by setting the environment variable `CUDA_VISIBLE_DEVICES`. Thus, you can keep the default GPU device id = 0 in RKeOps.

* Other runtime options, see `?runtime_options`

## Advanced use

### Precision

By default, RKeOps uses float 32bits precision for computations. Since R only considers 64bits floating point numbers, if you want to use float 32bits, input data and output results will be casted befors and after computations respectively in your RKeOps operator. If your application requires to use float 64bits (double) precision, keep in mind that you will suffer a performance loss (potentially not an issue on high-end GPUs). In any case, compensated summation reduction is available in KeOps to correct for 32bits floating point arithmetic errors.

### Data storage orientation

In R, matrices are stored using a column-major order, meaning that a $M \times D$ matrix is stored in memory as a succession of $D$ vectors of length $M$ representing each of its columns. A consequence is that two successive entries of a column are contiguous in memory, but two successive entries of a row are separated by $M$ elements. See this [page](https://en.wikipedia.org/wiki/Row-_and_column-major_order) for more details.<br><br>

For RKeOps to be computationnally efficient, it is important that elements of the input matrices are contiguous along the inner dimensions $D$ (or $D'$). Thus, it is recommended to use input matrices where the outer dimension (i.e. indexes $i$ or $j$) are the columns, and inner dimensions the rows, e.g. transpose matrices $\mathbf X^{t} = [x_{ki}]_{D \times M}$ or $\mathbf Y^{t} = [y_{k'i}]_{D' \times N}$.

> **_Important:_** In machine learning and statistics, we generally use data matrices where each sample/observation/individual is a row, i.e. matrices where the outer dimensions correspond to rows, e.g. $\mathbf X = [x_{ik}]_{M \times D}$, $\mathbf Y = [y_{ik'}]_{N \times D'}$.<br><br> This is the default using case of RKeOps. RKeOps will then automatically convert your matrices to their transpose, where the outer dimensions correspond to columns.<br><br> If you want to use data where the inner dimension directly corresponds to rows of your matrices, i.e. $\mathbf X^{t} = [x_{ki}]_{D \times M}$ or $\mathbf Y^{t} = [y_{k'i}]_{D' \times N}$, you just need to specify the input parameter `inner_dim=0` when calling your operator.

Example:
```{r storage_order, eval=FALSE}
# standard column reduction of a matrix product
op <- keops_kernel(formula = "Sum_Reduction((x|y), 1)",
                   args = c("x=Vi(3)", "y=Vj(3)"))

# data (inner dimension = columns)
nx <- 10
ny <- 15
# x_i = rows of the matrix X
X <- matrix(runif(nx*3), nrow=nx, ncol=3)
# y_j = rows of the matrix Y
Y <- matrix(runif(ny*3), nrow=ny, ncol=3)
# computing the result (here, by default `inner_dim=1` and columns corresponds
# to the inner dimension)
res <- op(list(X,Y))

# data (inner dimension = rows)
nx <- 10
ny <- 15
# x_i = columns of the matrix X
X <- matrix(runif(nx*3), nrow=3, ncol=nx)
# y_j = columns of the matrix Y
Y <- matrix(runif(ny*3), nrow=3, ncol=ny)
# computing the result (we specify `inner_dim=0` to indicate that rows
# corresponds to the inner dimension)
res <- op(list(X,Y), inner_dim=0)
```

### Compilation files and cleaning

The compilation of new operators produces shared library (or share object `.so`) files stored in a `build` sub-directory of the package installation directory, to be reused and avoid recompilation of already defined operators.

You can check where your compiled operators are stored by running `get_build_dir()`. To clean RKeOps install and remove all shared library files, you can run `clean_rkeops()`.

